---
title: "Median-odds-ratio"
author: "Steve Simon"
date: "May 10, 2017"
output: html_document
---

This program explores the median odds ratio, a measure of the size of a random effect in a random effects logistic regression model.

```{r preliminaries}
library(dplyr)
library(ggplot2)
library(magrittr)
```

This data set represents the number of cancer cells out of 400 that were killed by a radiation exposure. There were 27 experiments run, 3 per day across 9 different days.

```{r read-data}
f <- "http://www.statsci.org/data/general/radiatio.txt"
radiation <- read.table(file=f, header=TRUE)
radiation %<>% mutate(p=Survived/400) %>% mutate(label="Raw data")
print(radiation)
summary(radiation$p)
```

Let's plot the data, ignoring the day.

```{r plot, fig.width=7, fig.height=1}
radiation %>%
  ggplot(aes(label, p)) +
  geom_boxplot() +
  coord_flip()
```

It may not be obvious, but the data you are looking at is too variable to come from a binomial distribution.

```{r compare-to-simulation, fig.width=7, fig.height=1.33}
n <- dim(radiation)[1]
simulated_binomial <- rbinom(n, 400, mean(radiation$p))/400
tibble(p=simulated_binomial,
  label="Simulated binomial")                   %>%
  bind_rows(radiation)                          %>%
  ggplot(aes(label, p)) +
  geom_boxplot() +
  coord_flip()



```

There is significant day-to-day variation. How would you model that variation?

```{r glm}
fixed_model <- glm(p~1, data=radiation, family=binomial)
summary(fixed_model)
```